---
title: "Activity 2"
author: "Nick Malloy"
format: pdf
editor: visual
toc: True
---

# Exploratory Data Analysis 

```{r,echo = FALSE}
insuranceData <- read.csv("activity2_insurance.csv")
library(tidyverse)
library(ggplot2)
```

```{r,echo = FALSE}
insuranceData%>%
  select(age,smoker,charges)%>%
  ggplot(aes(x = age, y = charges))+
  geom_point(aes(color = smoker))
```
Here we can see some of the customers with the highest charge amounts are smokers and that age isn't a huge indicator of high charges.
```{r,echo = FALSE}
insuranceData%>%
  na.omit()%>%
  group_by(region)%>%
  summarise(meanCharges = mean(charges))
```
The Southeast region is showing the highest average charges overall, so there could be some geographical differences. 
```{r,echo = FALSE}
insuranceData%>%
  select(bmi,region,charges)%>%
  ggplot(aes(x=bmi, y = charges))+
  geom_point()+
  facet_wrap(~ region)
```
All regions follow the same sort of correlation for BMI and charges.
```{r,echo = FALSE}
insuranceData%>%
  select(bmi,children,charges)%>%
  ggplot(aes(x = bmi, y = charges))+
  geom_point(aes(color = "red"))+
  facet_wrap(~ children)
```
Customers with 3 or less children show the same general positive correlation between their BMI and charges. Also important to note that we don't have many customers with 4 or more children. 
```{r,echo = FALSE}
insuranceData2 <-
  insuranceData%>%
  select(sex,region,smoker)%>%
  group_by(sex,region)%>%
  mutate(TotalPeople = n())%>%
  pivot_longer(smoker, values_to = "smoker")%>%
  select(sex,region,smoker,TotalPeople)%>%
  group_by(sex,region,smoker)%>%
  summarise(PercentOfTotal = n()/TotalPeople)%>%
  distinct()%>%
  arrange(region)
split_insurance <- list()
num_rows <- nrow(insuranceData2)
for (i in seq(1, num_rows, by = 4)) {
  end_index <- min(i + 4 - 1, num_rows)
  current_split <- insuranceData2[i:end_index, ]
  split_insurance[[length(split_insurance) + 1]] <- current_split
}
split_insurance
```
Here are demographic data frames for each region, split into gender and whether they're a smoker or not. Percent of Total refers to their specific gender.